---
title: "Real-world Case Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-world Case Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(heteroTests)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates how to incorporate heteroTests into real analyses.
Each case study follows a workflow of model specification, diagnostic testing,
visualisation, and interpretation.

## Case study 1: Seismic station counts

R's built-in `quakes` dataset records 1,000 seismic events near Fiji. We
regress the number of reporting stations on event magnitude and depth, and
inspect the residual dispersion. The response is a count, so its variance is
expected to rise with its mean.

```{r}
quakes_model <- lm(stations ~ mag + depth, data = quakes)
summary(quakes_model)
```

Formal diagnostics flag heteroscedasticity consistent with the classical
literature on this dataset.

```{r}
white_q <- performWhiteTest(quakes_model, quakes)
bp_q <- performBPTest(quakes_model, quakes)
koenker_q <- performKoenkerTest(quakes_model, quakes)
list(White = white_q, Breusch_Pagan = bp_q, Koenker = koenker_q)
```

```{r fig.width=6, fig.height=4}
plot_data <- data.frame(
  fitted = fitted(quakes_model),
  residual = resid(quakes_model),
  mag = quakes$mag
)

ggplot(plot_data, aes(x = fitted, y = residual, colour = mag)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_c(option = "C") +
  geom_smooth(se = FALSE, colour = "black") +
  labs(
    x = "Fitted values",
    y = "Residuals",
    colour = "mag",
    title = "Heteroscedasticity in reported station counts"
  ) +
  theme_minimal()
```

*Interpretation.* All three tests reject homoskedasticity. The residual plot
reveals wider dispersion at higher predicted counts and for events of larger
`mag`. Weighted least squares is a common
remedy.

```{r}
wls_q <- fitWLS(quakes_model)
scale_comparison <- c(OLS = sigma(quakes_model), WLS = sigma(wls_q))
scale_comparison
```

The WLS fit reduces the residual scale, validating the remedy.

## Case study 2: Income volatility in simulated survey data

The `hetero_data` sample mimics survey responses with variance inflation driven
by the regressor `x`.

```{r}
data(hetero_data, package = "heteroTests")
survey_model <- lm(y ~ x, data = hetero_data)
summary(survey_model)
```

`HeteroDiagnostic()` orchestrates multiple tests and visualisations.

```{r}
survey_diag <- HeteroDiagnostic(survey_model, hetero_data)
test(survey_diag, tests = c("white", "breusch_pagan"))
performGlejserTest(survey_model, hetero_data, "x")
```

```{r fig.width=6, fig.height=4}
plot(survey_diag, plots = c("spread_level"))
```

*Interpretation.* Glejser's test and the scale-location plot both indicate that
variance increases with `x`. Depending on the modelling goal, analysts might log
transform the response or model the variance explicitly (e.g. via `lm` weights
or `glm` with a variance function).

## Case study 3: Engineering reliability assessment

The `diagnostic_data` dataset contains correlated predictors representing
engineering stress measures. Nonlinear effects and multicollinearity can mask
heteroscedasticity. We run the broader diagnostic suite to capture interactions
between variance and structure checks.

```{r}
data(diagnostic_data, package = "heteroTests")
diagnostic_ext <- transform(diagnostic_data, x1_sq = x1^2)
engineering_model <- lm(y ~ x1 + x1_sq + x2, data = diagnostic_ext)
engineer_results <- runDiagnostics(engineering_model, diagnostic_ext,
  tests = c("breusch_pagan", "koenker")
)
safe_htest <- function(fn, method) {
  tryCatch(
    fn(),
    error = function(e) {
      message(method, " unavailable: ", e$message)
      structure(
        list(
          statistic = c(statistic = NA_real_),
          parameter = NA_real_,
          p.value = NA_real_,
          method = paste0(method, " (failed)"),
          data.name = deparse(stats::formula(engineering_model))
        ),
        class = "htest"
      )
    }
  )
}
white_engineering <- safe_htest(
  function() performWhiteTest(engineering_model, diagnostic_ext, cross_products = FALSE),
  "White test"
)
bp_engineering <- safe_htest(
  function() performBPTest(engineering_model, diagnostic_ext),
  "Breusch-Pagan test"
)
koenker_engineering <- safe_htest(
  function() performKoenkerTest(engineering_model, diagnostic_ext),
  "Koenker test"
)
harvey_engineering <- safe_htest(
  function() performHarveyTest(engineering_model),
  "Harvey test"
)
engineer_tests <- list(
  white = white_engineering,
  breusch_pagan = bp_engineering,
  koenker = koenker_engineering,
  harvey = harvey_engineering
)
engineer_tests
```

The heteroscedasticity tests disagree, so we inspect the accompanying
multicollinearity and RESET diagnostics.

```{r}
engineer_results$vif
engineer_results$reset
```

Finally, we consolidate $p$-values to visualise which tests detect variance
instability.

```{r fig.width=6, fig.height=4}
pvals <- sapply(engineer_tests, function(x) x$p.value)
pval_df <- data.frame(
  test = names(pvals),
  p_value = as.numeric(pvals)
)

ggplot(pval_df, aes(x = reorder(test, p_value), y = p_value)) +
  geom_col(fill = "#0072B2", alpha = 0.8) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "#D55E00") +
  coord_cartesian(ylim = c(0, 1)) +
  coord_flip() +
  labs(
    x = "Test",
    y = "p-value",
    title = "Contrasting heteroscedasticity diagnostics"
  ) +
  theme_minimal()
```

*Interpretation.* Koenker's robust variant is conservative in this moderate
sample, whereas White's omnibus test remains sensitive to nonlinear variance.
The RESET test also signals functional-form misspecification, suggesting that a
variance stabilising transformation or spline terms could stabilise the error
variance while addressing nonlinearity.

## Summary

Across the three scenarios, heteroTests streamlines workflows by combining
statistical theory with practical diagnostics. The package encourages analysts to
inspect residual plots alongside formal tests and to validate remedies such as
weighted least squares or variance modelling.
