---
title: "heteroTests Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{heteroTests Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(heteroTests)
```

This vignette demonstrates the main workflow in greater depth. We use R's
built-in `quakes` dataset.

```{r}
head(quakes)
```

## 1. Fit an initial model

```{r}
base_model <- lm(stations ~ mag + depth, data = quakes)
summary(base_model)
```

## 2. Diagnose heteroscedasticity

```{r}
hd <- HeteroDiagnostic(base_model, quakes)
test(hd)
plot(hd)
```

## 3. Apply a remedy

```{r}
wls_model <- fitWLS(base_model)
summary(wls_model)
```

## 4. Compare models

```{r}
compareModelDiagnostics(list(base_model, wls_model))
```

The weighted least squares fit typically reduces the heteroscedasticity metrics.
On a weighted fit the tests are computed from the standardized residuals, the
raw residuals multiplied by the square root of their weights, so a drop in the
statistics means that the weighting worked.

## 5. Check the variance function

`fitWLS()` assumed that the log of the error variance is linear in the
regressors. `performVarianceFormTest()` tests that shape against the squares
and products of the same regressors.

```{r}
performVarianceFormTest(wls_model)
```

A small p-value says that the variance function needs those extra terms, and
that another `form` or `var_formula` in `fitWLS()` is worth trying. Unlike the
tests in step 4, this one allows for the weights having been estimated from
the same data.


