---
title: "heteroTests: A Unified and Reproducible Framework for Heteroscedasticity Diagnostics in R"
date: "2026-10-07"
abstract: >
  Heteroscedasticity diagnostics in R are distributed across multiple packages
  with different interfaces, return objects, assumptions, and workflow
  conventions. The heteroTests package provides a unified framework for
  diagnosing non-constant variance in linear-model workflows. It combines
  classical auxiliary-regression and group-wise procedures with rank-based,
  ARCH-type, resampling, robust, spatial, and scalable variants behind a
  consistent interface based on standard htest objects. This article describes
  the package architecture, validation strategy, batch and streaming workflows,
  and integration with common R modelling tools. Agreement with established
  reference implementations is used where available, while simulation studies
  assess empirical size and power under several variance structures and error
  distributions. Computational benchmarks evaluate scaling with sample size
  and illustrate the memory-bounded streaming implementations. A worked
  example shows how heteroTests connects diagnosis, interpretation, and
  remediation in a reproducible analysis.
draft: true
author:
  - name: Diogo Ribeiro
    affiliation: Faculty of Media Arts and Design, Technical University of Porto
    address:
      - Vila do Conde, Portugal
    url: https://orcid.org/0009-0001-2022-7072
    orcid: 0009-0001-2022-7072
    email: dfr@esmad.ipp.pt
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: heteroTests.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(heteroTests)
```

# Introduction

Heteroscedasticity is not a single diagnostic problem. Different procedures
target different alternatives, rely on different assumptions, and are useful
for different data structures. In R, this diversity is reflected in the
software ecosystem: established implementations are distributed across base R
and specialised packages such as \\CRANpkg{lmtest} and \\CRANpkg{car}
[@zeileis2002lmtest; @fox2019car]. That distribution is appropriate from a
methodological point of view, but it introduces a practical cost when several
diagnostics must be used in the same analysis. Function signatures differ,
return objects expose different fields, and model or data preparation often has
to be repeated before results can be compared or reported together.

The difficulty is therefore not the absence of individual tests. White's
general test [@white1980heteroskedasticity], the Breusch--Pagan test
[@breusch1979simple], Koenker's studentized variant [@koenker1981study], and
Cook--Weisberg diagnostics [@cook1983detection], among others, are well
established. The software problem is that a reproducible workflow frequently
needs more than one of them, together with a common representation of the model,
the data used for fitting, the diagnostic result, and the choices that determine
how the statistic was computed.

\\pkg{heteroTests} addresses this software problem. It provides individual
diagnostic functions, a registry that supports common orchestration, and a
higher-level execution layer that normalises model and data inputs before
running selected procedures. Results returned as `htest` objects are decorated
with package-specific metadata and a `hetero_test` class, while collections are
represented by `hetero_test_suite` or `hetero_grouped_suite` objects. This
keeps compatibility with familiar R hypothesis-test conventions while making
diagnostics easier to combine, tidy, inspect, and pass through automated
workflows.

The package does not propose that a single omnibus procedure should replace
method-specific judgement. Nor does it claim statistical novelty merely by
wrapping established tests. Its contribution is an execution and integration
layer around heterogeneous diagnostics: consistent inputs and outputs, a
registry for extension, batch execution, grouped-data handling, caching,
ecosystem adapters, and memory-bounded implementations of selected
auxiliary-regression tests.

A second design goal is to avoid making scalability a separate user workflow.
For supported diagnostics, `runHeteroTests()` estimates the size of the supplied
data and can switch automatically from the ordinary implementation to a
streaming implementation when a configurable memory threshold is exceeded.
The streaming procedures accumulate the cross-products required by the
auxiliary regression rather than materialising its full design matrix. This
preserves the statistic while reducing peak memory use.

The remainder of the article evaluates these design choices rather than
re-deriving the underlying tests. The next section describes the package
architecture and execution model. Later sections validate results against
reference implementations where possible, study empirical size and power under
controlled simulations, benchmark computational scaling, and demonstrate an
end-to-end diagnostic workflow.

# Package design

The package is organised in four layers: individual statistical procedures,
result normalisation, diagnostic orchestration, and ecosystem integration. This
separation allows a diagnostic to remain independently callable while also
participating in batch workflows.

## Individual diagnostics and result objects

Individual procedures retain explicit names such as `performWhiteTest()`,
`performBPTest()`, and `performKoenkerTest()`. The naming convention makes
the diagnostic operation visible while avoiding namespace collisions with
established functions in other packages.

Most hypothesis-test functions return standard R `htest` objects. When a
result is used through the orchestration layer, \\pkg{heteroTests} decorates it
with the `hetero_test` class and attaches metadata identifying the diagnostic,
the number of observations when available, the fitted model, the data source,
and execution-specific information. The original `htest` structure is
preserved, so conventional fields such as `statistic`, `parameter`,
`p.value`, and `method` remain available.

The additional class supports a common downstream interface. The package
implements `tidy()`, `glance()`, and `augment()` methods through
\\CRANpkg{generics}; suite-level methods combine individual results into data
frames suitable for reporting or subsequent analysis. This design avoids
requiring each underlying diagnostic to expose the same internal computation.

## Registry-based orchestration

The diagnostic registry maps character names to functions that accept a model
and data and return a diagnostic result. Built-in entries include the classical
auxiliary-regression tests together with resampling, rank-based,
high-dimensional, and spatial procedures. Users can extend the registry with
`registerDiagnostic()`, after which custom diagnostics can be selected by
`runHeteroTests()` in the same way as built-in entries.

This registry is deliberately simpler than a plugin framework. It provides a
stable dispatch point without requiring a new object hierarchy for every
method. The registry is also shared with the broader diagnostic machinery in
the package, which means that adding a compatible function is sufficient for it
to participate in batch execution.

The default call

```{r basic-suite, eval=FALSE}
fit <- lm(mpg ~ wt + qsec, data = mtcars)

runHeteroTests(fit, mtcars)
```

runs White and Breusch--Pagan diagnostics. Additional registered procedures are
selected explicitly:

```{r selected-suite, eval=FALSE}
runHeteroTests(
  fit,
  mtcars,
  tests = c("white", "breusch_pagan", "koenker", "wild_bootstrap")
)
```

A small default set is intentional. Running many diagnostics automatically would
blur the distinction between exploratory screening and a test chosen because
its assumptions and alternative hypothesis are appropriate for the model.

## Input normalisation and grouped analyses

The public execution layer accepts fitted `lm` and `glm` objects, formulas,
selected \\CRANpkg{tidymodels} workflow or parsnip objects, and compatible
data-frame representations. Internally, these inputs are normalised before a
diagnostic is dispatched. Formula inputs are fitted using the supplied data,
while supported workflow and parsnip objects are reduced to their underlying
engine fits.

Grouped tibbles receive different treatment. When a formula-based specification
can be recovered, the model is refitted independently within each group and the
result is returned as a `hetero_grouped_suite`, together with the group keys.
The package therefore treats grouping as a set of separate modelling problems,
rather than applying one global residual vector to pre-partitioned data.

The input layer also accepts `data.table` and `dtplyr`-backed data and
provides dedicated support for survey designs. These adapters are not intended
to redefine the statistical meaning of the underlying diagnostics; they reduce
the amount of conversion code required before a supported model can enter the
same diagnostic pipeline.

## Caching and failure handling

Repeated diagnostic calls can reuse cached results. `runHeteroTests()` builds
an analysis key from the model, data, selected tests, and relevant execution
parameters when caching is enabled and the optional hashing dependency is
available. Cached objects are returned using the same suite classes as newly
computed results.

Errors are analysed at the orchestration layer rather than being silently
discarded. The package records diagnostic context, can attempt a supported
fallback, and produces structured failure results for grouped analyses where
continuing with the remaining groups is preferable to aborting the entire
operation. For ordinary ungrouped calls, unrecoverable errors are still raised
to the caller.

## Memory-bounded auxiliary regressions

White, Breusch--Pagan, and Koenker diagnostics also have streaming
implementations. Their common computational problem is an auxiliary regression
whose statistic depends on its coefficient of determination. Rather than
constructing the complete auxiliary design matrix, the streaming code processes
observations in chunks and accumulates

\[
  X^\top X, \qquad X^\top y, \qquad
  \sum_i y_i, \qquad \text{and} \qquad \sum_i y_i^2.
\]

These quantities are sufficient to solve the auxiliary least-squares problem
and recover the residual and total sums of squares. If
\(\widehat{\beta}=(X^\top X)^{-1}X^\top y\), the residual sum of squares can
be written as

\[
  \mathrm{SSE}=y^\top y-\widehat{\beta}^\top X^\top y,
\]

while the total sum of squares is obtained from the accumulated first and
second moments of the response. The resulting \(R^2\) is therefore computed
without retaining all rows of the auxiliary design simultaneously.

This matters most for White's test, where squares and cross-products can expand
the auxiliary design substantially. The implementation also checks for
rank-deficient accumulated systems and can omit interaction expansion when the
number of regressors exceeds a configurable limit. If \\CRANpkg{Matrix} is
available, sufficiently large chunk cross-products may be evaluated using sparse
matrices.

`runHeteroTests()` estimates the size of the supplied data and uses these
streaming implementations automatically for supported diagnostics when the
configured threshold is exceeded. The threshold and chunk size remain user
controls, so the same API can be used for both ordinary in-memory and
memory-constrained analyses.

# Validation strategy

Validation has two complementary layers. Where an established implementation
exists, heteroTests is compared against reference packages such as
\\CRANpkg{lmtest} and \\CRANpkg{car}. Procedures without a suitable reference
implementation are checked against their published definitions and tested for
expected statistical behaviour.

The final article will report the package-version-specific validation results
generated from the reproducibility scripts rather than copying numerical claims
from earlier drafts.

# Simulation study

The simulation study will evaluate empirical size and power under multiple
sample sizes, variance functions, and error distributions. All tables and
figures in this section will be generated from R code used by the manuscript.

# Computational performance

Benchmarks will measure runtime and memory behaviour as sample size increases,
with particular attention to the streaming implementations. Results will be
regenerated for the release analysed in this article.

# A reproducible diagnostic workflow

This section will present one end-to-end analysis: fit a model, diagnose
non-constant variance, interpret conflicting diagnostics, apply an appropriate
remediation, and compare the resulting model diagnostics.

# Comparison with existing R software

The comparison will focus on software scope and workflow differences rather
than treating existing packages as competitors. In particular, the article
will distinguish specialised reference implementations from the unified
interface and orchestration provided by heteroTests.

# Discussion

No single heteroscedasticity diagnostic is uniformly preferable. Test choice
depends on the suspected variance structure, sample size, distributional
assumptions, dependence structure, and the inferential purpose of the model.
The package therefore exposes multiple procedures while retaining the
information needed to understand how each result was obtained.

# Summary

heteroTests provides a common software layer for heteroscedasticity diagnostics
in R. Its contribution is not a replacement for the underlying statistical
literature, but a reproducible and extensible framework that makes heterogeneous
diagnostics easier to validate, combine, scale, and integrate into modelling
workflows.
