---
title: "heteroTests: A Unified and Reproducible Framework for Heteroscedasticity Diagnostics in R"
date: "2026-10-08"
abstract: >
  Heteroscedasticity diagnostics in R are distributed across multiple packages
  with different interfaces, return objects, assumptions, and workflow
  conventions. The heteroTests package provides a unified framework for
  diagnosing non-constant variance in linear-model workflows. It combines
  classical auxiliary-regression and group-wise procedures with rank-based,
  ARCH-type, resampling, robust, spatial, and scalable variants behind a
  consistent interface based on standard htest objects. This article describes
  the package architecture, validation strategy, batch and streaming workflows,
  and integration with common R modelling tools. Agreement with established
  reference implementations is used where available, while simulation studies
  assess empirical size and power under several variance structures and error
  distributions. Computational benchmarks evaluate scaling with sample size
  and illustrate the memory-bounded streaming implementations. A worked
  example shows how heteroTests connects diagnosis, interpretation, and
  remediation in a reproducible analysis, and how the variance model behind
  the remediation can itself be tested.
draft: false
author:
  - name: Diogo Ribeiro
    affiliation: Faculty of Media Arts and Design, Technical University of Porto
    address:
      - Vila do Conde, Portugal
    url: https://orcid.org/0009-0001-2022-7072
    orcid: 0009-0001-2022-7072
    email: dfr@esmad.ipp.pt
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: heteroTests.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

# RJournal.sty loads inconsolata, which loads xkeyval; with xkeyval present a
# comma inside \includegraphics[alt={...}] breaks the figure and prints the alt
# text instead. pdfLaTeX does not use alt text, so keep it for HTML only.
if (knitr::is_latex_output()) {
  # The rjtools LaTeX template does not scale figures to the text width.
  knitr::opts_chunk$set(out.width = "\\linewidth")
  knitr::opts_hooks$set(fig.alt = function(options) {
    options$fig.alt <- NULL
    options
  })
}

library(heteroTests)

sim_size <- utils::read.csv(
  file.path("generated", "simulation_size_table.csv"),
  check.names = FALSE
)
sim_power <- utils::read.csv(
  file.path("generated", "simulation_power_full_strength.csv"),
  stringsAsFactors = FALSE
)
perf_speed <- utils::read.csv(
  file.path("generated", "performance_reference_speed.csv"),
  stringsAsFactors = FALSE
)
perf_accuracy <- utils::read.csv(
  file.path("generated", "performance_accuracy_table.csv"),
  stringsAsFactors = FALSE
)
perf_stream <- utils::read.csv(
  file.path("generated", "performance_streaming_summary.csv"),
  stringsAsFactors = FALSE
)

# Rejection rates for weighted fits come from the validation study that is
# installed with the package, so they describe the version being rendered.
weighted_path <- system.file(
  "validation", "weighted-fits-size.csv",
  package = "heteroTests"
)
if (!nzchar(weighted_path)) {
  stop(
    "This article needs heteroTests >= 0.12.0, which installs the ",
    "weighted-fit validation results.",
    call. = FALSE
  )
}
weighted_fits <- utils::read.csv(weighted_path, stringsAsFactors = FALSE)
weighted_known <- subset(
  weighted_fits,
  weights == "known" & !is.na(rejection_rate)
)
weighted_rate <- function(test_name, block, n_obs) {
  rows <- weighted_fits$test == test_name &
    weighted_fits$weights == block &
    weighted_fits$n == n_obs &
    weighted_fits$errors == "normal"
  weighted_fits$rejection_rate[rows]
}

pct <- function(x, digits = 1L) {
  sprintf(paste0("%.", digits, "f\\%%"), 100 * x)
}
```

# Introduction

Heteroscedasticity is not a single diagnostic problem. Different procedures
target different alternatives, rely on different assumptions, and are useful
for different data structures. In R, this diversity is reflected in the
software ecosystem: established implementations are distributed across base R
and specialised packages such as \CRANpkg{lmtest} and \CRANpkg{car}
[@zeileis2002lmtest; @fox2019car]. That distribution is appropriate from a
methodological point of view, but it introduces a practical cost when several
diagnostics must be used in the same analysis. Function signatures differ,
return objects expose different fields, and model or data preparation often has
to be repeated before results can be compared or reported together.

The difficulty is therefore not the absence of individual tests. White's
general test [@white1980heteroskedasticity], the Breusch--Pagan test
[@breusch1979simple], Koenker's studentized variant [@koenker1981study], and
Cook--Weisberg diagnostics [@cook1983detection], among others, are well
established. The software problem is that a reproducible workflow frequently
needs more than one of them, together with a common representation of the model,
the data used for fitting, the diagnostic result, and the choices that determine
how the statistic was computed.

\CRANpkg{heteroTests} addresses this software problem. It provides individual
diagnostic functions, a registry that supports common orchestration, and a
higher-level execution layer that normalises model and data inputs before
running selected procedures. Results returned as `htest` objects are decorated
with package-specific metadata and a `hetero_test` class, while collections are
represented by `hetero_test_suite` or `hetero_grouped_suite` objects. This
keeps compatibility with familiar R hypothesis-test conventions while making
diagnostics easier to combine, tidy, inspect, and pass through automated
workflows.

The package does not propose that a single omnibus procedure should replace
method-specific judgement. Nor does it claim statistical novelty merely by
wrapping established tests. Its contribution is an execution and integration
layer around heterogeneous diagnostics: consistent inputs and outputs, a
registry for extension, batch execution, grouped-data handling, caching,
ecosystem adapters, and memory-bounded implementations of selected
auxiliary-regression tests.

A second design goal is to avoid making scalability a separate user workflow.
For supported diagnostics, `runHeteroTests()` estimates the size of the supplied
data and can switch automatically from the ordinary implementation to a
streaming implementation when a configurable memory threshold is exceeded.
The streaming procedures accumulate the cross-products required by the
auxiliary regression rather than materialising its full design matrix. This
preserves the statistic while reducing peak memory use.

A third is to keep the diagnostics meaningful after a model has been adjusted
for heteroscedasticity. Tests applied to a weighted fit are computed on the
weighted scale, where they ask whether the weights are adequate, and the
variance function used to construct feasible weights can be tested as a null
hypothesis in its own right.

The remainder of the article evaluates these design choices rather than
re-deriving the underlying tests. The next section describes the package
architecture and execution model. Later sections validate results against
reference implementations where possible, study empirical size and power under
controlled simulations, benchmark computational scaling, and demonstrate an
end-to-end diagnostic workflow.

# Package design

The package is organised in four layers: individual statistical procedures,
result normalisation, diagnostic orchestration, and ecosystem integration. This
separation allows a diagnostic to remain independently callable while also
participating in batch workflows.

## Individual diagnostics and result objects

Individual procedures retain explicit names such as `performWhiteTest()`,
`performBPTest()`, and `performKoenkerTest()`. The naming convention makes
the diagnostic operation visible while avoiding namespace collisions with
established functions in other packages.

Most hypothesis-test functions return standard R `htest` objects. When a
result is used through the orchestration layer, \CRANpkg{heteroTests} decorates it
with the `hetero_test` class and attaches metadata identifying the diagnostic,
the number of observations when available, the fitted model, the data source,
and execution-specific information. The original `htest` structure is
preserved, so conventional fields such as `statistic`, `parameter`,
`p.value`, and `method` remain available.

The additional class supports a common downstream interface. The package
implements `tidy()`, `glance()`, and `augment()` methods through
\CRANpkg{generics}; suite-level methods combine individual results into data
frames suitable for reporting or subsequent analysis. This design avoids
requiring each underlying diagnostic to expose the same internal computation.

## Registry-based orchestration

The diagnostic registry maps character names to functions that accept a model
and data and return a diagnostic result. Built-in entries include the classical
auxiliary-regression tests together with resampling, rank-based,
high-dimensional, and spatial procedures. Users can extend the registry with
`registerDiagnostic()`, after which custom diagnostics can be selected by
`runHeteroTests()` in the same way as built-in entries.

This registry is deliberately simpler than a plugin framework. It provides a
stable dispatch point without requiring a new object hierarchy for every
method. The registry is also shared with the broader diagnostic machinery in
the package, which means that adding a compatible function is sufficient for it
to participate in batch execution.

The default call

```{r basic-suite, eval=FALSE}
fit <- lm(mpg ~ wt + qsec, data = mtcars)

runHeteroTests(fit, mtcars)
```

runs White and Breusch--Pagan diagnostics. Additional registered procedures are
selected explicitly:

```{r selected-suite, eval=FALSE}
runHeteroTests(
  fit,
  mtcars,
  tests = c("white", "breusch_pagan", "koenker", "wild_bootstrap")
)
```

A small default set is intentional. Running many diagnostics automatically would
blur the distinction between exploratory screening and a test chosen because
its assumptions and alternative hypothesis are appropriate for the model.

## Input normalisation and grouped analyses

The public execution layer accepts fitted `lm` and `glm` objects, formulas,
selected \CRANpkg{tidymodels} workflow or parsnip objects, and compatible
data-frame representations. Internally, these inputs are normalised before a
diagnostic is dispatched. Formula inputs are fitted using the supplied data,
while supported workflow and parsnip objects are reduced to their underlying
engine fits.

Grouped tibbles receive different treatment. When a formula-based specification
can be recovered, the model is refitted independently within each group and the
result is returned as a `hetero_grouped_suite`, together with the group keys.
The package therefore treats grouping as a set of separate modelling problems,
rather than applying one global residual vector to pre-partitioned data.

The input layer also accepts `data.table` and `dtplyr`-backed data and
provides dedicated support for survey designs. These adapters are not intended
to redefine the statistical meaning of the underlying diagnostics; they reduce
the amount of conversion code required before a supported model can enter the
same diagnostic pipeline.

## Caching and failure handling

Repeated diagnostic calls can reuse cached results. `runHeteroTests()` builds
an analysis key from the model, data, selected tests, and relevant execution
parameters when caching is enabled and the optional hashing dependency is
available. Cached objects are returned using the same suite classes as newly
computed results.

Errors are analysed at the orchestration layer rather than being silently
discarded. The package records diagnostic context, can attempt a supported
fallback, and produces structured failure results for grouped analyses where
continuing with the remaining groups is preferable to aborting the entire
operation. For ordinary ungrouped calls, unrecoverable errors are still raised
to the caller.

## Memory-bounded auxiliary regressions

White, Breusch--Pagan, and Koenker diagnostics also have streaming
implementations. Their common computational problem is an auxiliary regression
whose statistic depends on its coefficient of determination. Rather than
constructing the complete auxiliary design matrix, the streaming code processes
observations in chunks and accumulates

\[
  X^\top X, \qquad X^\top y, \qquad
  \sum_i y_i, \qquad \text{and} \qquad \sum_i y_i^2.
\]

These quantities are sufficient to solve the auxiliary least-squares problem
and recover the residual and total sums of squares. If
\(\widehat{\beta}=(X^\top X)^{-1}X^\top y\), the residual sum of squares can
be written as

\[
  \mathrm{SSE}=y^\top y-\widehat{\beta}^\top X^\top y,
\]

while the total sum of squares is obtained from the accumulated first and
second moments of the response. The resulting \(R^2\) is therefore computed
without retaining all rows of the auxiliary design simultaneously.

This matters most for White's test, where squares and cross-products can expand
the auxiliary design substantially. The implementation also checks for
rank-deficient accumulated systems and can omit interaction expansion when the
number of regressors exceeds a configurable limit. If \CRANpkg{Matrix} is
available, sufficiently large chunk cross-products may be evaluated using sparse
matrices.

`runHeteroTests()` estimates the size of the supplied data and uses these
streaming implementations automatically for supported diagnostics when the
configured threshold is exceeded. The threshold and chunk size remain user
controls, so the same API can be used for both ordinary in-memory and
memory-constrained analyses.

## Weighted fits and variance functions

Two design decisions concern models that have already been adjusted for
heteroscedasticity. The first is what a diagnostic should compute for a fit
obtained with `lm(..., weights = w)`. Such a fit asserts that the error
variance is proportional to \(1/w_i\), so its raw residuals are
heteroscedastic by assumption and a statistic computed from them cannot show
whether the weights are adequate. For weighted fits the diagnostics therefore
use the Pearson residuals, the raw residuals multiplied by the square roots of
their weights. These are the residuals of the equivalent unweighted regression
and have constant variance when the weights are correct, so the null
hypothesis becomes that the weights are adequate. Variance regressors,
grouping variables, and ordering variables remain on their original scale.
Procedures that regenerate the response and refit the mean model, such as the
resampling tests, would discard the weights; for a weighted fit they stop with
an error instead of testing a different model.

The second decision is to make the variance model used for remediation
explicit and testable. `fitWLS()` estimates a variance function from the OLS
residuals and refits by feasible weighted least squares. Its `var_formula`
argument names the variance regressors, and `form` selects an exponential
function, in which the log variance is linear in those regressors
[@harvey1976estimating], or a power function, in which it is linear in their
logarithms [@park1966estimation]. The fitted function is stored with the
returned model.

`performVarianceFormTest()` takes such a function as its null hypothesis. The
other diagnostics test constant variance, so a rejection shows that the
variance changes with some variables but not whether a particular functional
form describes the change. The variance-function test regresses the squared
standardised residuals of the weighted fit on the variance regressors and on
additional terms, by default the squares and cross-products of those
regressors, and reports the \(F\) test of the additional terms. It is a
regression-based specification test
[@wooldridge1990unified; @wooldridge1991application]. Retaining the variance
regressors in the auxiliary regression removes the first-order effect of
having estimated the variance function from the same residuals.

# Validation strategy

The validation strategy is organised around the claim being checked rather than
around the package function that happens to implement it. Three forms of
evidence are used: agreement with an established implementation, independent
reconstruction of a published statistic, and behavioural tests of statistical
or computational properties.

## Agreement with established implementations

Where another R implementation exposes the same statistic under the same
assumptions, the package test suite compares both the statistic and the
associated p-value. This is the strongest validation available for wrappers of
well-established methods because it checks not only formula transcription but
also practical details such as degrees of freedom, alternative hypotheses, and
centering conventions.

The current suite includes direct comparisons with \CRANpkg{lmtest} for the
classical Breusch--Pagan, studentized Breusch--Pagan/Koenker, and
Goldfeld--Quandt tests [@goldfeld1965tests; @zeileis2002lmtest]. Mean-centred Levene,
median-centred Brown--Forsythe, and Cook--Weisberg/non-constant-variance
diagnostics are checked against \CRANpkg{car} [@fox2019car]. Group-wise
variance procedures are also compared with base R implementations for Bartlett
and Fligner--Killeen tests and, where available, with \CRANpkg{vartest} for
Hartley and O'Brien procedures.

Weighted fits show why the choice of reference matters. For a model fitted
with weights, `car::ncvTest()` computes its score statistic from Pearson
residuals, whereas `lmtest::bptest()` retains the raw residuals and weights
the auxiliary regression. The two functions therefore test different
hypotheses, and only the first asks whether the weights are adequate.
\\CRANpkg{heteroTests} follows the Pearson-residual convention described
earlier. On weighted fits its score test is compared with \\CRANpkg{car}, and
its Goldfeld--Quandt test with `lmtest::gqtest()` applied to the equivalent
transformed regression.

These comparisons use explicit numerical tolerances in the automated test
suite. The tolerance depends on the implementation pair: some functions match
to floating-point equality, while others are checked within a small numerical
tolerance because the two implementations may differ in intermediate linear
algebra without changing the intended statistic.

## Independent reconstruction

An external package is not always available, and using another implementation
is not sufficient when both could share the same mistake. For this reason, a
second group of tests reconstructs the claimed statistic directly from its
published definition or from lower-level base-R operations.

Examples include the Cook--Weisberg score statistic [@cook1983detection],
Harvey's multiplicative-variance LM statistic [@harvey1976estimating],
Szroeter's ordered-variance statistic, Engle's ARCH LM statistic
[@engle1982autoregressive], and the McLeod--Li procedure [@mcleod1983diagnosis], which is checked
against a Ljung--Box calculation on squared residuals. The same approach is
used for O'Brien's transformed observations and for degrees-of-freedom
behaviour under rank-deficient auxiliary models.

This layer is particularly useful for implementation details that are easy to
get subtly wrong. Examples caught during package development include scaling of
ordered-variance statistics, auxiliary-model degrees of freedom, and the
construction of resampling null distributions.

## Behavioural and invariant checks

Some properties are better validated as invariants than by comparing a single
number. Property-based tests verify that supported diagnostics return valid
`htest` objects with finite statistics and p-values in \([0,1]\) over randomly
generated inputs. Simulation-based regression tests check that selected
classical procedures have empirical Type-I error close to the nominal level,
using a tolerance based on the Monte Carlo standard error rather than a fixed
arbitrary margin.

The same principle covers weighted fits. When heteroscedastic data are fitted
with the correct weights, a diagnostic should reject at its nominal rate. This
property has to be tested by simulation, because no comparison on unweighted
fits can reveal it: before version 0.12.0 the heteroscedasticity diagnostics
read the raw residuals of a weighted fit, agreed with their references on
every unweighted comparison, and rejected a correctly weighted model in every
replication of the package's validation design. In the current version the
`r nrow(weighted_known)` procedures that accept a weighted fit reject between
`r pct(min(weighted_known$rejection_rate), 0)` and
`r pct(max(weighted_known$rejection_rate), 0)` of such models at the 5% level,
based on `r weighted_known$replications[1]` replications each.

Resampling procedures receive separate checks because an apparently plausible
implementation can still produce invalid p-values. Bootstrap and permutation
p-values use the finite-simulation convention

\[
  p = \frac{1 + \#\{T_b \ge T_{\mathrm{obs}}\}}
           {B_{\mathrm{eff}} + 1},
\]

which prevents an estimated p-value of exactly zero and accounts for failed
replicates through the effective number of simulations. Null-imposed residual
resampling is also checked for leverage correction, while unsupported model
classes, weighted fits, and transformed responses are rejected rather than
silently refitted under a different statistical model.

Finally, the streaming implementations are compared directly with their
ordinary in-memory counterparts. For White, Breusch--Pagan, and Koenker tests,
the test suite checks equality of the statistic and p-value within numerical
tolerance for identical data. This verifies the principal claim of the
streaming layer: reduced peak memory use should not define a different
diagnostic.

## Reproducibility of the evidence

The article does not hard-code validation numbers. Reference comparisons and
behavioural checks live in the package test suite, while the empirical size,
power, and performance results reported later are read at render time from
recorded runs, each identified by package version, environment, and seed. This
separation is intentional. Unit and reference tests answer whether an
implementation is computing the intended object; simulation and benchmark
sections answer how those procedures behave under selected data-generating
processes and computational loads.

The distinction also limits the claims made here. Agreement with another
implementation is evidence of computational equivalence for the cases tested,
not proof that the underlying statistical procedure is appropriate for every
analysis. Likewise, satisfactory empirical size under a simulation design does
not establish finite-sample validity outside that design. The purpose of this
validation framework is therefore narrower: to make the software behaviour
auditable and to reduce the risk that integration work changes the statistic a
diagnostic is supposed to compute.

# Simulation study

The simulation study separates calibration under the null from power under
specific alternatives. This distinction is important because a procedure that
rejects frequently under a heteroscedastic alternative is not useful if it also
substantially exceeds its nominal rejection probability under homoscedasticity.

All experiments use the linear mean model

\[
  y_i = 1 + 2x_i + \sigma(x_i)\varepsilon_i,
  \qquad x_i \sim U(0,1),
\]

with independent errors. The significance level is \(\alpha=0.05\). Within
each Monte Carlo replication the same simulated data set is supplied to all
diagnostics, so differences in rejection rates are not inflated by using
different random samples for different tests.

The study includes White, classical Breusch--Pagan, Koenker, and a two-sided
Goldfeld--Quandt test ordered by \(x\). These procedures represent different
ways of detecting variance structure: general auxiliary-regression expansion,
linear score-type alternatives, a studentized variant less dependent on
normality, and an ordered two-sample variance comparison.

## Empirical size

Empirical size is estimated under \(\sigma(x)=1\) for sample sizes
\(n\in\{50,100,250\}\). Two error distributions are considered: Gaussian
errors and Student \(t_5\) errors rescaled to unit variance. The heavier-tailed
case is included because agreement with a reference implementation does not
imply that an asymptotic approximation is well calibrated under non-normal
finite samples.

For each reported rejection rate \(\widehat p\), the simulation records the
Monte Carlo standard error

\[
  \operatorname{MCSE}(\widehat p)
  = \sqrt{\frac{\widehat p(1-\widehat p)}{B_{\mathrm{eff}}}},
\]

where \(B_{\mathrm{eff}}\) is the number of valid replications. Approximate
95% Monte Carlo intervals are reported as
\(\widehat p \pm 1.96\operatorname{MCSE}(\widehat p)\), truncated to
\([0,1]\). These intervals quantify simulation uncertainty; they are not
confidence intervals for a population parameter estimated from observed data.

## Power alternatives

Power is studied under three variance patterns with Gaussian errors:

- **linear:** variance scale increases monotonically with \(x\);
- **step:** the variance changes abruptly at \(x=0.5\);
- **U-shaped:** variance is smallest near the centre of the covariate range and
  increases toward both extremes.

For each pattern, an effect-strength parameter
\(\gamma\in\{0.25,0.50,0.75,1.00\}\) interpolates between constant
variance and the target profile. The resulting standard-deviation vector is
renormalised to unit root-mean-square scale. Consequently, increasing
\(\gamma\) changes the *shape* of heteroscedasticity without simply changing
the overall noise magnitude. This avoids the common error of calling a global
multiplicative scale change an increase in heteroscedastic effect size.

The non-monotone U-shaped alternative is included deliberately. A test ordered
by a single covariate direction can be powerful against monotone changes while
having much less sensitivity to variance patterns that increase at both ends of
the covariate range. The purpose of the study is therefore comparative rather
than to identify a universally dominant diagnostic.

## Reproducible experiment

The complete experiment is implemented in
`paper/scripts/simulation_study.R`. The manuscript run uses an explicit seed and
writes both results and simulation metadata to `paper/generated/`:

```{r simulation-command, eval=FALSE}
system2(
  "Rscript",
  c("paper/scripts/simulation_study.R", "2000")
)
```

The default manuscript configuration uses 2,000 replications per design cell.
With a rejection probability near 0.05 this corresponds to a Monte Carlo
standard error of approximately 0.005, which is small enough to distinguish
material size distortions from ordinary simulation noise without pretending
that the estimated rejection rates are exact.

Numerical results and figures are intentionally read from committed extracts of
the successful workflow artifact rather than copied from earlier drafts. The
full experiment used `heteroTests` 0.11.2 under R 4.6.1 on Ubuntu 24.04.5;
the workflow and artifact identifiers are recorded in
`paper/generated/PROVENANCE.md`. Version 0.12.0, which the rest of the article
describes, changes these four procedures only for weighted fits. The study
fits unweighted models, and a rerun of the full experiment under 0.12.0
reproduced every rejection rate in those extracts exactly.

## Simulation results

Under Gaussian errors, all four procedures remain close to the nominal 5%
level across the three sample sizes. At \(n=100\), for example, rejection
rates range from
`r sprintf("%.2f\\%%", min(subset(sim_size, Error == "gaussian" & n == 100)[["Rejection (%)"]]))`
to
`r sprintf("%.2f\\%%", max(subset(sim_size, Error == "gaussian" & n == 100)[["Rejection (%)"]]))`.
The heavier-tailed null separates the procedures much more clearly. At
\(n=250\), the classical Breusch--Pagan test rejects
`r sprintf("%.1f\\%%", subset(sim_size, Error == "t5" & n == 250 & Test == "Breusch-Pagan")[["Rejection (%)"]])`
of homoscedastic samples and Goldfeld--Quandt rejects
`r sprintf("%.1f\\%%", subset(sim_size, Error == "t5" & n == 250 & Test == "Goldfeld-Quandt")[["Rejection (%)"]])`,
whereas Koenker and White reject
`r sprintf("%.1f\\%%", subset(sim_size, Error == "t5" & n == 250 & Test == "Koenker")[["Rejection (%)"]])`
and
`r sprintf("%.1f\\%%", subset(sim_size, Error == "t5" & n == 250 & Test == "White")[["Rejection (%)"]])`,
respectively. Thus the simulation reproduces the practical reason for keeping
the studentized and more general procedures distinct from their
normality-sensitive counterparts.

Table \@ref(tab:size-results) reports the middle sample size for both error
distributions; the complete size grid is stored in the generated CSV.

```{r size-results, echo=FALSE}
size_n100 <- subset(sim_size, n == 100)
size_n100[["Rejection (%)"]] <- round(size_n100[["Rejection (%)"]], 2)
size_n100[["MC lower (%)"]] <- round(size_n100[["MC lower (%)"]], 2)
size_n100[["MC upper (%)"]] <- round(size_n100[["MC upper (%)"]], 2)

knitr::kable(
  size_n100[, c("Error", "Test", "Rejection (%)", "MC lower (%)", "MC upper (%)")],
  # kable() does not escape captions; a bare % would start a LaTeX comment.
  caption = paste0(
    "Empirical size at n = 100 with 2,000 Monte Carlo replications. ",
    "Bounds are approximate 95",
    if (knitr::is_latex_output()) "\\%" else "%",
    " Monte Carlo intervals."
  ),
  booktabs = TRUE
)
```

Power depends strongly on the variance shape. At full effect strength and
\(n=100\), all four procedures have high power against the monotone linear
alternative: the observed rejection rates range from
`r pct(min(subset(sim_power, pattern == "Linear" & n == 100)$estimate), 1)`
to
`r pct(max(subset(sim_power, pattern == "Linear" & n == 100)$estimate), 1)`.
The step alternative is similarly easy at this sample size. The U-shaped
alternative is qualitatively different: White reaches
`r pct(subset(sim_power, pattern == "U-shaped" & n == 100 & test == "White")$estimate, 1)`,
while Breusch--Pagan, Koenker, and Goldfeld--Quandt reach only
`r pct(subset(sim_power, pattern == "U-shaped" & n == 100 & test == "Breusch-Pagan")$estimate, 1)`,
`r pct(subset(sim_power, pattern == "U-shaped" & n == 100 & test == "Koenker")$estimate, 1)`,
and
`r pct(subset(sim_power, pattern == "U-shaped" & n == 100 & test == "Goldfeld-Quandt")$estimate, 1)`.
This is not a ranking of the procedures in general; it illustrates that tests
targeting monotone or ordered variance changes can have little power against a
symmetric non-monotone alternative.

```{r power-full-strength, echo=FALSE, fig.width=7.5, fig.height=4.3, fig.cap="Empirical rejection rates at full heteroscedastic effect strength across variance patterns and sample sizes. Each point is based on 2,000 Monte Carlo replications.", fig.alt="Three-panel line chart of empirical rejection rates for linear, step, and U-shaped variance patterns. Lines compare White, Breusch-Pagan, Koenker, and Goldfeld-Quandt diagnostics across sample sizes 50, 100, and 250."}
power_plot <- sim_power
power_plot$n <- factor(power_plot$n, levels = c(50, 100, 250))

ggplot2::ggplot(
  power_plot,
  ggplot2::aes(
    x = n,
    y = estimate,
    group = test,
    linetype = test,
    shape = test
  )
) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 1.8) +
  ggplot2::facet_wrap(~pattern, nrow = 1) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    labels = function(x) paste0(round(100 * x), "%")
  ) +
  ggplot2::labs(
    x = "Sample size",
    y = "Empirical rejection rate",
    linetype = "Diagnostic",
    shape = "Diagnostic"
  ) +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::theme(legend.position = "bottom")
```

# Computational performance

The performance study is designed around two separate questions. The first is
the cost of the package's validation and integration layer relative to
established reference implementations. The second is whether the streaming
implementations provide a useful memory--time trade-off while preserving the
underlying statistic. Combining these questions into one benchmark would be
misleading because the reference packages and the streaming code solve
different software problems.

## Reference-package overhead

For statistically equivalent procedures, \CRANpkg{heteroTests} is benchmarked
against the corresponding implementation in \CRANpkg{lmtest} or
\CRANpkg{car}. The benchmark covers classical Breusch--Pagan, Koenker's
studentized form, and the Cook--Weisberg/non-constant-variance score test.

Synthetic regression data are generated with four predictors and several
variance patterns. Sample size increases from small inputs to \(10^5\)
observations in the manuscript configuration. For each test and sample size we
record elapsed time and, when the \CRANpkg{bench} package is available,
allocated memory. The benchmark also records absolute differences in the
reported statistic and p-value relative to the reference implementation.

Raw wall-clock ratios are interpreted cautiously. At small sample sizes both
implementations may complete near the timer's resolution, so a ratio can become
unstable or infinite even when the absolute difference is negligible. The
article therefore reports absolute timings and statistical agreement rather
than presenting a single cross-package "speed-up" number.

## Exact versus streaming implementations

White, Breusch--Pagan, and Koenker tests are benchmarked separately using their
ordinary and streaming implementations on the same generated datasets. The
streaming variants process fixed-size chunks and accumulate the cross-products
described earlier, while the exact variants follow the ordinary in-memory path.

For every benchmark cell, the experiment records

- elapsed time;
- allocated memory when available;
- the absolute difference in the test statistic; and
- the absolute difference in the p-value.

The statistic and p-value differences are the primary correctness checks. A
streaming implementation is only useful if reduced memory requirements do not
change the diagnostic being computed. Runtime is a secondary trade-off: chunked
processing may be slower for data that already fit comfortably in memory.

## Reproducible benchmark workflow

The complete benchmark is implemented in `paper/scripts/performance_study.R`; summary
tables and figures are generated by `paper/scripts/summarise_performance.R`. Both
scripts write to `paper/generated/`, together with metadata describing seeds,
sample sizes, replication counts, chunk size, and whether memory profiling was
available.

```{r performance-command, eval=FALSE}
system2("Rscript", "paper/scripts/performance_study.R")
system2("Rscript", "paper/scripts/summarise_performance.R")
```

The manuscript does not use the retired `scaling_plot.pdf` as evidence. The
reported performance results were generated from `heteroTests` 0.11.2 under
R 4.6.1 on Ubuntu 24.04.5 and are preserved with workflow provenance. They
were not regenerated for version 0.12.0, whose changes to the benchmarked
procedures are confined to the handling of weighted fits.

## Performance results

The reference comparisons show that integration and validation have a measurable
runtime cost, but do not change the reported inference. Across all benchmarked
sample sizes, the largest median p-value difference from the corresponding
\CRANpkg{lmtest} or \CRANpkg{car} implementation is
`r format(max(perf_accuracy$median_p_value_diff, na.rm = TRUE), scientific = TRUE, digits = 3)`.
At \(n=10^5\), median elapsed times for \CRANpkg{heteroTests} are
`r sprintf("%.3f", subset(perf_speed, test == "breusch_pagan" & package == "heteroTests" & sample_size == 100000)$median_time)` s
for Breusch--Pagan,
`r sprintf("%.3f", subset(perf_speed, test == "koenker" & package == "heteroTests" & sample_size == 100000)$median_time)` s
for Koenker, and
`r sprintf("%.3f", subset(perf_speed, test == "ncv" & package == "heteroTests" & sample_size == 100000)$median_time)` s
for NCV. The corresponding reference timings are retained in the generated
benchmark table rather than collapsed into a single speed ratio.

The streaming comparison gives a more direct software result. At
\(n=250{,}000\), measured allocation falls from
`r sprintf("%.1f", subset(perf_stream, test == "breusch_pagan" & sample_size == 250000 & implementation == "exact")$memory_mb)` MB
to
`r sprintf("%.1f", subset(perf_stream, test == "breusch_pagan" & sample_size == 250000 & implementation == "streaming")$memory_mb)` MB
for Breusch--Pagan, from
`r sprintf("%.1f", subset(perf_stream, test == "koenker" & sample_size == 250000 & implementation == "exact")$memory_mb)` MB
to
`r sprintf("%.1f", subset(perf_stream, test == "koenker" & sample_size == 250000 & implementation == "streaming")$memory_mb)` MB
for Koenker, and from
`r sprintf("%.1f", subset(perf_stream, test == "white" & sample_size == 250000 & implementation == "exact")$memory_mb)` MB
to
`r sprintf("%.1f", subset(perf_stream, test == "white" & sample_size == 250000 & implementation == "streaming")$memory_mb)` MB
for White. In this benchmark the chunked implementations are also faster, though
that should be treated as an empirical result of this design rather than a
guarantee.

Most importantly, the largest absolute statistic difference between exact and
streaming computations at \(n=250{,}000\) is
`r format(max(subset(perf_stream, sample_size == 250000)$statistic_diff), scientific = TRUE, digits = 3)`,
and the largest p-value difference is
`r format(max(subset(perf_stream, sample_size == 250000)$p_value_diff), scientific = TRUE, digits = 3)`.
The benchmark therefore supports the intended interpretation of streaming as a
different computational path to essentially the same statistic, not as an
approximate test.

```{r streaming-memory, echo=FALSE, fig.width=7.2, fig.height=4.3, fig.cap="Allocated memory for exact and streaming implementations. Values are medians from three benchmark replications with chunk size 5,000.", fig.alt="Three-panel line chart comparing allocated memory for exact and streaming Breusch-Pagan, Koenker, and White implementations. Streaming uses less memory at each displayed sample size."}
stream_mem <- subset(perf_stream, is.finite(memory_mb))
stream_mem$sample_size <- factor(
  stream_mem$sample_size,
  levels = c(1000, 10000, 100000, 250000)
)

ggplot2::ggplot(
  stream_mem,
  ggplot2::aes(
    x = sample_size,
    y = memory_mb,
    group = implementation,
    linetype = implementation,
    shape = implementation
  )
) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 1.8) +
  ggplot2::facet_wrap(~test, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Sample size",
    y = "Allocated memory (MB, log scale)",
    linetype = "Implementation",
    shape = "Implementation"
  ) +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::theme(legend.position = "bottom")
```

# A reproducible diagnostic workflow

The previous sections treat diagnostics as statistical and computational
objects. This section shows how the package is intended to be used in an
analysis: diagnose a specific model, interpret the evidence, choose a
follow-up whose assumptions match the variance pattern, and check the variance
model that the follow-up relies on.

A controlled example is useful here because the correct mean and variance
structures are known. We generate

\[
  y_i = 1 + 2x_i + \varepsilon_i,
  \qquad
  \varepsilon_i \sim N\!\left(0, \exp(1.8x_i)\right),
\]

with \(x_i\sim U(-1,1)\). The mean model is therefore correctly specified as
linear, while the log variance is linear in the same predictor. This is a case
where feasible weighted least squares is an interpretable follow-up rather than
a generic attempt to make a diagnostic p-value disappear.

```{r workflow-data}
set.seed(20261008)

n_workflow <- 500L
workflow_data <- data.frame(
  x = stats::runif(n_workflow, -1, 1)
)
workflow_data$sigma <- exp(0.9 * workflow_data$x)
workflow_data$y <- 1 + 2 * workflow_data$x +
  stats::rnorm(n_workflow, sd = workflow_data$sigma)

ols_fit <- stats::lm(y ~ x, data = workflow_data)
```

## Diagnose the fitted model

The example uses three regression-based diagnostics with complementary
properties: White's general test, classical Breusch--Pagan, and Koenker's
studentized form. They are selected explicitly rather than by running every
available diagnostic.

```{r workflow-diagnostics}
workflow_tests <- runHeteroTests(
  ols_fit,
  workflow_data,
  tests = c("white", "breusch_pagan", "koenker"),
  use_cache = FALSE,
  progress = FALSE
)

workflow_table <- data.frame(
  Diagnostic = c("White", "Breusch-Pagan", "Koenker"),
  Statistic = vapply(
    workflow_tests,
    function(x) unname(x$statistic[[1]]),
    numeric(1)
  ),
  p.value = vapply(
    workflow_tests,
    function(x) x$p.value,
    numeric(1)
  ),
  row.names = NULL
)

workflow_table$Statistic <- round(workflow_table$Statistic, 3)
workflow_table$p.value <- format.pval(
  workflow_table$p.value,
  digits = 3,
  eps = 0.001
)

knitr::kable(
  workflow_table,
  caption = "Heteroscedasticity diagnostics for the initial OLS model.",
  booktabs = TRUE
)
```

The diagnostics can also be passed to the package's lightweight recommendation
layer:

```{r workflow-remediation-suggestion}
workflow_suggestion <- suggestRemediation(workflow_tests)
workflow_suggestion
```

The recommendation should be read as workflow guidance, not as an automatic
model-selection rule. A significant test says that the constant-variance
assumption is not supported for the fitted mean model; it does not by itself
choose between a variance model, heteroscedasticity-consistent covariance
estimation, a transformation, or a different mean specification.

## Model the variance

Here the data-generating mechanism makes a variance model appropriate. The
`fitWLS()` helper estimates a feasible-GLS model by regressing logged squared
OLS residuals on the original design matrix, exponentiating the fitted
log-variance, and using its inverse as the weight. It does not invert individual
squared residuals. This is the default, exponential variance function; the
`var_formula` and `form` arguments select other variance regressors or a power
function.

```{r workflow-wls}
wls_fit <- fitWLS(ols_fit)

coef_summary <- function(model, method) {
  tab <- summary(model)$coefficients
  data.frame(
    Method = method,
    Term = rownames(tab),
    Estimate = tab[, "Estimate"],
    Std.Error = tab[, "Std. Error"],
    row.names = NULL
  )
}

coefficient_comparison <- rbind(
  coef_summary(ols_fit, "OLS"),
  coef_summary(wls_fit, "Feasible WLS")
)

coefficient_comparison$Estimate <- round(
  coefficient_comparison$Estimate,
  3
)
coefficient_comparison$Std.Error <- round(
  coefficient_comparison$Std.Error,
  3
)

knitr::kable(
  coefficient_comparison,
  caption = paste(
    "Coefficient estimates and model-based standard errors before and after",
    "feasible weighted least squares."
  ),
  booktabs = TRUE
)
```

Because the mean model is unchanged, the point of WLS is not to force the raw
residuals to have constant spread. If the fitted variance is
\(\widehat\sigma_i^2\), the relevant residual scale is

\[
  r_i^{(w)} =
  \frac{y_i-\widehat y_i}{\widehat\sigma_i}
  =
  \sqrt{w_i}\,(y_i-\widehat y_i).
\]

A useful before/after check therefore compares the OLS residuals with these
weighted residuals, not the raw residuals from the weighted fit.

```{r workflow-weighted-residuals, echo=FALSE, fig.width=7, fig.height=3.8, fig.cap="Residual spread before and after variance modelling. The WLS panel shows residuals on the weighted scale.", fig.alt="Two residual scatterplots against x. The OLS panel shows increasing residual spread, while the weighted WLS panel shows a more stable spread across x."}
workflow_residuals <- rbind(
  data.frame(
    x = workflow_data$x,
    residual = stats::residuals(ols_fit),
    Model = "OLS residuals"
  ),
  data.frame(
    x = workflow_data$x,
    residual = sqrt(stats::weights(wls_fit)) * stats::residuals(wls_fit),
    Model = "WLS weighted residuals"
  )
)

ggplot2::ggplot(
  workflow_residuals,
  ggplot2::aes(x = x, y = residual)
) +
  ggplot2::geom_point(alpha = 0.35, size = 0.8) +
  ggplot2::geom_smooth(method = "loess", se = FALSE) +
  ggplot2::facet_wrap(~Model, nrow = 1) +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed") +
  ggplot2::labs(
    x = "x",
    y = "Residual"
  ) +
  ggplot2::theme_minimal(base_size = 9)
```

For a compact numerical summary, the correlation between residual magnitude and
the predictor changes from
`r sprintf("%.3f", cor(abs(residuals(ols_fit)), workflow_data$x))`
for OLS to
`r sprintf("%.3f", cor(abs(sqrt(weights(wls_fit)) * residuals(wls_fit)), workflow_data$x))`
on the weighted WLS scale. This is only a descriptive check, but it shows what
the remediation is intended to accomplish: account for a systematic variance
pattern while retaining the same linear mean specification.

## Check the variance model

The weighted scale is also the one the package's diagnostics use for a
weighted fit, so the same suite can be applied to `wls_fit` directly. Table
\@ref(tab:workflow-retest-table) sets the resulting statistics beside those of
the OLS fit.

```{r workflow-retest}
wls_tests <- runHeteroTests(
  wls_fit,
  workflow_data,
  tests = c("white", "breusch_pagan", "koenker"),
  use_cache = FALSE,
  progress = FALSE
)
```

```{r workflow-retest-table, echo=FALSE}
suite_statistics <- function(suite) {
  vapply(suite, function(x) unname(x$statistic[[1]]), numeric(1))
}

retest_table <- data.frame(
  Diagnostic = c("White", "Breusch-Pagan", "Koenker"),
  OLS = round(suite_statistics(workflow_tests), 3),
  WLS = round(suite_statistics(wls_tests), 3)
)

knitr::kable(
  retest_table,
  caption = "Diagnostic statistics for the OLS fit and for the feasible WLS fit, the latter computed from weighted residuals.",
  booktabs = TRUE,
  row.names = FALSE,
  col.names = c("Diagnostic", "OLS fit", "Feasible WLS fit")
)
```

All three statistics fall by more than two orders of magnitude. They are
useful for this comparison, but their p-values should not be read as
calibrated tests of the weighted model. The reference distributions treat the
weights as known, whereas `fitWLS()` estimated them from the same residuals.
In the validation study installed with the package, with a correctly specified
variance function and Gaussian errors, the Koenker test applied to a
`fitWLS()` fit rejected
`r pct(weighted_rate("performKoenkerTest", "estimated", 150), 0)`
of the time at the 5% level with 150 observations and
`r pct(weighted_rate("performKoenkerTest", "estimated", 600), 0)`
with 600, so the discrepancy does not shrink as the sample grows. Harvey's
test, which repeats the regression the weights were estimated from, rejected
`r pct(weighted_rate("performHarveyTest", "estimated", 600), 0)`
of the time with 600 observations.

`performVarianceFormTest()` is designed for this situation. Its null
hypothesis is the variance function that produced the weights, and in the same
validation designs it rejected
`r pct(weighted_rate("performVarianceFormTest", "estimated", 150), 0)`
of the time with 150 observations and
`r pct(weighted_rate("performVarianceFormTest", "estimated", 600), 0)`
with 600. Applied to `wls_fit`, it tests the fitted exponential function
against the square of the variance regressor. Two further calls test a
deliberately incomplete variance function, which contains only a quadratic
term, first against its default added term and then against the linear term it
omits.

```{r workflow-variance-form}
variance_check <- performVarianceFormTest(wls_fit)

quadratic_default <- performVarianceFormTest(
  ols_fit,
  var_formula = ~ I(x^2)
)
quadratic_against_x <- performVarianceFormTest(
  ols_fit,
  var_formula = ~ I(x^2),
  against = ~ x
)
```

```{r workflow-variance-form-table, echo=FALSE}
form_row <- function(result, variance, added) {
  data.frame(
    Variance = variance,
    Added = added,
    F = round(unname(result$statistic), 3),
    p.value = format.pval(result$p.value, digits = 3, eps = 0.001)
  )
}

form_table <- rbind(
  form_row(variance_check, "x", "x squared"),
  form_row(quadratic_default, "x squared", "x to the fourth power"),
  form_row(quadratic_against_x, "x squared", "x")
)

knitr::kable(
  form_table,
  caption = "Variance-function tests for the worked example. Each row tests a log-variance that is linear in the listed regressor against the listed added term.",
  booktabs = TRUE,
  row.names = FALSE,
  col.names = c("Variance regressor", "Added term", "F", "p-value")
)
```

Table \@ref(tab:workflow-variance-form-table) reports the three tests. The
first row is the function `fitWLS()` fitted. The test finds no evidence
against it, and the estimated slope of the log variance is
`r sprintf("%.2f", variance_check$estimate[["x"]])`
against a true value of 1.8. The other two rows show what such a result does
and does not establish. The quadratic-only function omits the linear
dependence that generated the data, yet it is not rejected when tested against
its default added term, because the misfit does not lie in that direction.
Tested against the omitted linear term, it is rejected decisively. The test has power against the terms it is given; a
non-rejection is evidence that those terms are not needed, not that the
functional form is correct.

A different inferential goal would lead to a different follow-up. If the
coefficient estimates were the primary target and no variance model were
scientifically justified, heteroscedasticity-consistent covariance estimation
would be a natural alternative [@zeileis2006sandwich]. Conversely, evidence of
mean-model misspecification should be addressed before treating the variance
alone. The diagnostic layer is therefore a decision point in the modelling
workflow, not an instruction to apply WLS after every rejected null.

# Comparison with existing R software

The package is best understood as an integration layer around a heterogeneous
set of diagnostics, not as a replacement for the specialised packages from
which several reference implementations are drawn. Table
\@ref(tab:software-comparison) summarises the roles most relevant to the
workflow in this article.

```{r software-comparison, echo=FALSE}
software_comparison <- data.frame(
  Package = c(
    "stats",
    "lmtest",
    "car",
    "vartest",
    "sandwich",
    "heteroTests"
  ),
  Primary.role = c(
    "General statistical infrastructure",
    "Regression specification and diagnostic tests",
    "Regression diagnostics and companion methods",
    "Tests for equality of variances",
    "Robust covariance estimation",
    "Heteroscedasticity diagnostic orchestration"
  ),
  Examples = c(
    "Bartlett; Fligner-Killeen",
    "Breusch-Pagan; Goldfeld-Quandt",
    "NCV/Cook-Weisberg; Levene",
    "Hartley; O'Brien",
    "HC covariance estimators",
    "Unified tests; suites; streaming"
  ),
  Workflow.contribution = c(
    "Reference implementations for classical group tests",
    "Established regression-test implementations",
    "Established regression and group diagnostics",
    "Specialised group-variance procedures",
    "Post-diagnostic robust inference",
    "Common API, metadata, adapters and batch execution"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  software_comparison,
  caption = "Roles of selected R packages used directly or as reference points in the heteroTests workflow.",
  booktabs = TRUE,
  col.names = c(
    "Package",
    "Primary role",
    "Examples",
    "Role in this workflow"
  )
)
```

## Established diagnostic implementations

Base R already provides high-quality procedures for several classical
equal-variance problems, including Bartlett and Fligner--Killeen tests.
\CRANpkg{heteroTests} does not improve these statistics by reimplementing them;
the value of including compatible procedures is that their results can
participate in the same reporting and orchestration layer as
regression-specific diagnostics.

\CRANpkg{lmtest} provides established regression diagnostic procedures,
including Breusch--Pagan and Goldfeld--Quandt tests
[@zeileis2002lmtest]. These implementations are used as numerical reference
points in the validation suite. When an analyst needs only one such procedure,
calling `lmtest::bptest()` or `lmtest::gqtest()` directly is simpler than
introducing another abstraction layer.

\CRANpkg{car} has a broader regression-diagnostics role and provides, among
other tools, the non-constant-variance score test and Levene-type variance
comparisons [@fox2019car]. These functions also serve as reference
implementations in the package tests. The relationship is therefore
complementary: \CRANpkg{car} supplies mature specialised diagnostics, while
\CRANpkg{heteroTests} supplies a common execution and result layer when these
kinds of diagnostics must be combined with others.

\CRANpkg{vartest} concentrates on tests for equality of variances. The
\CRANpkg{heteroTests} validation suite compares its Hartley and O'Brien
implementations against \CRANpkg{vartest} where that package is available. For an
analysis centred specifically on group-variance testing, the specialised
package remains a natural direct interface.

## Detection and robust inference are different tasks

\CRANpkg{sandwich} addresses a different problem. Its
heteroscedasticity-consistent covariance estimators modify coefficient
inference without requiring the analyst to specify a parametric variance model
[@zeileis2006sandwich]. A diagnostic package and a robust-covariance package
therefore answer different questions: one provides evidence about the
constant-variance assumption, while the other provides an inferential strategy
that can remain valid when that assumption is relaxed.

This distinction matters in practice. The worked example used feasible WLS
because the variance structure was deliberately modelled. If the coefficient
estimates were the primary target and no scientifically defensible variance
model were available, a \CRANpkg{sandwich} covariance estimator could be a more
appropriate follow-up. \CRANpkg{heteroTests} should therefore be viewed as
complementary to robust covariance estimation rather than as a substitute for
it.

## What the integration layer adds

The software contribution of \CRANpkg{heteroTests} appears when an analysis
requires several diagnostics or must be embedded in a larger reproducible
workflow. In that setting the package adds five pieces of infrastructure that
are not properties of any individual heteroscedasticity test:

1. **Common result semantics.** Hypothesis-test outputs retain the familiar
   `htest` structure and can be decorated as `hetero_test` objects with
   diagnostic and execution metadata.
2. **Registry-based orchestration.** `runHeteroTests()` selects registered
   diagnostics by name, supports user-defined entries, and returns suite objects
   with a common downstream interface.
3. **Input adapters.** Supported formulas, fitted models, grouped tibbles,
   `data.table`/dtplyr objects, survey designs, and selected tidymodels objects
   enter the same preprocessing path.
4. **Reproducibility and validation.** Reference comparisons, behavioural tests,
   explicit failure handling, and common metadata reduce the amount of bridging
   code needed when several diagnostics are reported together.
5. **Alternative computational paths.** Selected auxiliary-regression tests can
   switch to memory-bounded streaming implementations without changing the
   public diagnostic interface.

These conveniences have a cost. The performance study shows measurable
overhead relative to a direct call to a narrowly focused reference function,
particularly for Breusch--Pagan and Koenker diagnostics. That overhead reflects
input validation, result decoration, and the surrounding orchestration layer;
it should not be presented as evidence that the statistical procedure itself
is computationally superior.

The practical boundary is therefore straightforward. For a single established
test, the specialised implementation is often the shortest and clearest
choice. \CRANpkg{heteroTests} becomes useful when the analysis needs multiple
diagnostic families, consistent result handling, extensible orchestration,
large-data execution paths, or a reproducible bridge from diagnosis to
subsequent modelling decisions.

# Discussion

The results in this article reinforce a point that is easy to lose when many
diagnostics are available behind one interface: heteroscedasticity testing is
not a model-selection tournament. Different procedures are sensitive to
different departures from constant variance, and their finite-sample behaviour
depends on assumptions that are not erased by software unification.

The simulation study illustrates this directly. Under Gaussian homoscedastic
errors, the four procedures considered remain close to the nominal 5% level,
but their behaviour separates under heavier tails. Classical Breusch--Pagan and
Goldfeld--Quandt tests become substantially oversized under standardized
\(t_5\) errors, whereas Koenker and White remain much closer to nominal size.
The distinction between classical and studentized score tests is therefore not
an implementation detail. It affects the practical meaning of the result when
normality is questionable.

Power is equally alternative-specific. Monotone linear and step changes are
detected readily by several procedures in the simulation, while the symmetric
U-shaped alternative strongly favours White's more general auxiliary expansion.
This does not establish White as uniformly preferable; it shows why a test
chosen for an ordered or monotone variance alternative can have little power
when the actual variance structure is non-monotone. A useful workflow therefore
starts from a plausible variance mechanism, not from the desire to run the
largest possible collection of tests.

That point also limits how `runHeteroTests()` should be interpreted. The
function is an orchestration mechanism, not a multiple-testing procedure.
Running many correlated diagnostics and then reporting the smallest p-value
would alter the error rate and is not justified by the package interface. Batch
execution is most useful when a small set of procedures has been selected for
different but scientifically defensible alternatives, or when the exercise is
explicitly exploratory and reported as such.

## Statistical scope of the software adapters

The input-normalisation layer broadens software compatibility, not statistical
validity. Accepting a `glm`, a survey design, grouped data, or a tidymodels
object does not imply that every registered diagnostic is theoretically valid
for every such model. Each procedure retains the assumptions of its underlying
statistical definition. The adapters reduce conversion and bookkeeping code;
they do not convert a linear-model diagnostic into a general theorem for other
model classes.

The same distinction applies to grouped execution. Refitting a formula within
each group produces a reproducible set of separate diagnostics, but inference
across those groups still raises the ordinary issues of multiplicity,
dependence, and unequal sample size. The suite object records and organises the
results; it does not solve those inferential problems automatically.

## Computational trade-offs

The streaming implementation reduces peak allocation by replacing the full
auxiliary design with accumulated cross-products. The benchmark shows that this
can materially reduce measured memory use while preserving the reported
statistic to numerical precision in the designs studied. However, this
computational route is not free of trade-offs.

The calculation works with \(X^\top X\), whose condition number is the square of
the condition number of \(X\). Ill-conditioned auxiliary designs can therefore
be more numerically delicate than a decomposition applied directly to the full
design matrix. The implementation checks for rank deficiency and falls back to
more stable solves where possible, but the streaming path should still be
interpreted as a memory-oriented algorithm, not as a universally superior
linear-algebra method.

White's test introduces an additional dimensionality issue because squares and
cross-products expand the auxiliary design as the predictor count increases.
The implementation limits interaction expansion beyond a configurable number of
regressors. That safeguard prevents uncontrolled growth in memory and
computation; it changes the auxiliary specification and should therefore be
treated as an explicit computational choice rather than as an exact substitute
for the unrestricted White expansion.

The performance comparison with \CRANpkg{lmtest} and \CRANpkg{car} also
shows the cost of abstraction. Input checking, result decoration, and
orchestration add runtime relative to direct calls for some diagnostics. At the
same time, the numerical agreement results indicate that the additional
software layer is not changing the intended inference in the reference cases
tested. The relevant trade-off is therefore convenience and reproducibility
against modest overhead, not statistical quality against speed.

## Remediation is a modelling decision

Detecting non-constant variance does not determine the next model. The worked
example uses feasible WLS because the simulated variance is log-linear in the
model design, matching the variance model estimated by `fitWLS()`. In an
observational analysis, that structure would need scientific or empirical
justification. `performVarianceFormTest()` supplies part of the empirical
side by testing a fitted variance function against added terms, but it cannot
confirm a functional form, and variance functions that are close over the
observed range of the regressors will all be accepted. Diagnostics of
constant variance applied to a `fitWLS()` fit remain informative for
comparison with the original model; they are not calibrated tests, because
they treat estimated weights as known.

Other goals lead to different responses. Heteroscedasticity-consistent
covariance estimation can preserve the original coefficient estimates while
changing their estimated uncertainty. A transformation changes the target model
itself. Explicit generalized least squares models the variance structure.
Misspecification of the conditional mean should be addressed before attributing
all residual pattern to the variance. The package exposes several follow-up
helpers, but none removes the analyst's responsibility to decide which
inferential problem is actually being solved.

## Limits of the empirical evidence

The simulation study covers one linear mean model, three variance patterns,
three sample sizes, two null error distributions, and four diagnostics. Its
findings describe those designs; they are not an exhaustive power ranking. In
particular, panel, spatial, clustered, high-dimensional, and time-series
settings require separate evaluation. The rejection rates quoted for weighted
fits and for the variance-function test come from the package's own validation
study, which uses a different design, and are not part of that comparison.

The performance study is likewise platform-specific. The reported timings and
allocations come from one R version and one Linux Actions environment, with a
fixed predictor dimension and benchmark configuration. Absolute runtime should
not be interpreted as a hardware-independent property of the package. The more
portable findings are the observed numerical agreement with reference
implementations and the relative reduction in allocation between exact and
streaming paths on the same platform.

Finally, the package's breadth creates a maintenance obligation. A unified
interface is useful only if the individual procedures remain tied to their
published definitions and reference implementations. The validation matrix,
simulation tests, and explicit provenance used in this article are therefore
part of the software contribution rather than auxiliary documentation.

# Conclusion

\CRANpkg{heteroTests} provides a common software layer for heteroscedasticity
diagnostics in R. Its contribution is not a new omnibus test and not a claim
that one diagnostic should replace established specialised implementations.
Instead, the package standardises execution and result handling, supports
extensible batch workflows, records diagnostic context, validates selected
implementations against independent references, and supplies memory-bounded
paths for large auxiliary regressions. It also follows an analysis past the
first rejection: diagnostics of a weighted fit are computed on the weighted
scale, and the variance function behind the weights can itself be tested.

The empirical results support that narrower claim. Reference comparisons agree
to numerical precision for the procedures evaluated. The simulation study
shows why method choice still matters: classical Breusch--Pagan and
Goldfeld--Quandt tests are sensitive to heavy-tailed null errors in the studied
design, while White's broader auxiliary specification is much more sensitive to
a non-monotone U-shaped variance alternative. The performance study shows that
streaming White, Breusch--Pagan, and Koenker calculations can reduce measured
allocation substantially while reproducing the in-memory statistics to
numerical precision on the benchmark platform.

The practical value is therefore reproducibility of the *workflow* rather than
automatic selection of the *method*. Analysts can combine several justified
variance diagnostics without writing bespoke bridging code, preserve the
information needed to reconstruct what was computed, and choose a follow-up
model or covariance strategy based on the scientific purpose of the analysis.
