---
title: "heteroTests: A Unified and Reproducible Framework for Heteroscedasticity Diagnostics in R"
date: "2026-10-07"
abstract: >
  Heteroscedasticity diagnostics in R are distributed across multiple packages
  with different interfaces, return objects, assumptions, and workflow
  conventions. The heteroTests package provides a unified framework for
  diagnosing non-constant variance in linear-model workflows. It combines
  classical auxiliary-regression and group-wise procedures with rank-based,
  ARCH-type, resampling, robust, spatial, and scalable variants behind a
  consistent interface based on standard htest objects. This article describes
  the package architecture, validation strategy, batch and streaming workflows,
  and integration with common R modelling tools. Agreement with established
  reference implementations is used where available, while simulation studies
  assess empirical size and power under several variance structures and error
  distributions. Computational benchmarks evaluate scaling with sample size
  and illustrate the memory-bounded streaming implementations. A worked
  example shows how heteroTests connects diagnosis, interpretation, and
  remediation in a reproducible analysis.
draft: true
author:
  - name: Diogo Ribeiro
    affiliation: Faculty of Media Arts and Design, Technical University of Porto
    address:
      - Vila do Conde, Portugal
    url: https://orcid.org/0009-0001-2022-7072
    orcid: 0009-0001-2022-7072
    email: dfr@esmad.ipp.pt
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: heteroTests.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(heteroTests)
```

# Introduction

Heteroscedasticity is not a single diagnostic problem. Different procedures
target different alternatives, rely on different assumptions, and are useful
for different data structures. In R, this diversity is reflected in the
software ecosystem: established implementations are distributed across base R
and specialised packages such as \\CRANpkg{lmtest} and \\CRANpkg{car}
[@zeileis2002lmtest; @fox2019car]. That distribution is appropriate from a
methodological point of view, but it introduces a practical cost when several
diagnostics must be used in the same analysis. Function signatures differ,
return objects expose different fields, and model or data preparation often has
to be repeated before results can be compared or reported together.

The difficulty is therefore not the absence of individual tests. White's
general test [@white1980heteroskedasticity], the Breusch--Pagan test
[@breusch1979simple], Koenker's studentized variant [@koenker1981study], and
Cook--Weisberg diagnostics [@cook1983detection], among others, are well
established. The software problem is that a reproducible workflow frequently
needs more than one of them, together with a common representation of the model,
the data used for fitting, the diagnostic result, and the choices that determine
how the statistic was computed.

\\pkg{heteroTests} addresses this software problem. It provides individual
diagnostic functions, a registry that supports common orchestration, and a
higher-level execution layer that normalises model and data inputs before
running selected procedures. Results returned as `htest` objects are decorated
with package-specific metadata and a `hetero_test` class, while collections are
represented by `hetero_test_suite` or `hetero_grouped_suite` objects. This
keeps compatibility with familiar R hypothesis-test conventions while making
diagnostics easier to combine, tidy, inspect, and pass through automated
workflows.

The package does not propose that a single omnibus procedure should replace
method-specific judgement. Nor does it claim statistical novelty merely by
wrapping established tests. Its contribution is an execution and integration
layer around heterogeneous diagnostics: consistent inputs and outputs, a
registry for extension, batch execution, grouped-data handling, caching,
ecosystem adapters, and memory-bounded implementations of selected
auxiliary-regression tests.

A second design goal is to avoid making scalability a separate user workflow.
For supported diagnostics, `runHeteroTests()` estimates the size of the supplied
data and can switch automatically from the ordinary implementation to a
streaming implementation when a configurable memory threshold is exceeded.
The streaming procedures accumulate the cross-products required by the
auxiliary regression rather than materialising its full design matrix. This
preserves the statistic while reducing peak memory use.

The remainder of the article evaluates these design choices rather than
re-deriving the underlying tests. The next section describes the package
architecture and execution model. Later sections validate results against
reference implementations where possible, study empirical size and power under
controlled simulations, benchmark computational scaling, and demonstrate an
end-to-end diagnostic workflow.

# Package design

The package is organised in four layers: individual statistical procedures,
result normalisation, diagnostic orchestration, and ecosystem integration. This
separation allows a diagnostic to remain independently callable while also
participating in batch workflows.

## Individual diagnostics and result objects

Individual procedures retain explicit names such as `performWhiteTest()`,
`performBPTest()`, and `performKoenkerTest()`. The naming convention makes
the diagnostic operation visible while avoiding namespace collisions with
established functions in other packages.

Most hypothesis-test functions return standard R `htest` objects. When a
result is used through the orchestration layer, \\pkg{heteroTests} decorates it
with the `hetero_test` class and attaches metadata identifying the diagnostic,
the number of observations when available, the fitted model, the data source,
and execution-specific information. The original `htest` structure is
preserved, so conventional fields such as `statistic`, `parameter`,
`p.value`, and `method` remain available.

The additional class supports a common downstream interface. The package
implements `tidy()`, `glance()`, and `augment()` methods through
\\CRANpkg{generics}; suite-level methods combine individual results into data
frames suitable for reporting or subsequent analysis. This design avoids
requiring each underlying diagnostic to expose the same internal computation.

## Registry-based orchestration

The diagnostic registry maps character names to functions that accept a model
and data and return a diagnostic result. Built-in entries include the classical
auxiliary-regression tests together with resampling, rank-based,
high-dimensional, and spatial procedures. Users can extend the registry with
`registerDiagnostic()`, after which custom diagnostics can be selected by
`runHeteroTests()` in the same way as built-in entries.

This registry is deliberately simpler than a plugin framework. It provides a
stable dispatch point without requiring a new object hierarchy for every
method. The registry is also shared with the broader diagnostic machinery in
the package, which means that adding a compatible function is sufficient for it
to participate in batch execution.

The default call

```{r basic-suite, eval=FALSE}
fit <- lm(mpg ~ wt + qsec, data = mtcars)

runHeteroTests(fit, mtcars)
```

runs White and Breusch--Pagan diagnostics. Additional registered procedures are
selected explicitly:

```{r selected-suite, eval=FALSE}
runHeteroTests(
  fit,
  mtcars,
  tests = c("white", "breusch_pagan", "koenker", "wild_bootstrap")
)
```

A small default set is intentional. Running many diagnostics automatically would
blur the distinction between exploratory screening and a test chosen because
its assumptions and alternative hypothesis are appropriate for the model.

## Input normalisation and grouped analyses

The public execution layer accepts fitted `lm` and `glm` objects, formulas,
selected \\CRANpkg{tidymodels} workflow or parsnip objects, and compatible
data-frame representations. Internally, these inputs are normalised before a
diagnostic is dispatched. Formula inputs are fitted using the supplied data,
while supported workflow and parsnip objects are reduced to their underlying
engine fits.

Grouped tibbles receive different treatment. When a formula-based specification
can be recovered, the model is refitted independently within each group and the
result is returned as a `hetero_grouped_suite`, together with the group keys.
The package therefore treats grouping as a set of separate modelling problems,
rather than applying one global residual vector to pre-partitioned data.

The input layer also accepts `data.table` and `dtplyr`-backed data and
provides dedicated support for survey designs. These adapters are not intended
to redefine the statistical meaning of the underlying diagnostics; they reduce
the amount of conversion code required before a supported model can enter the
same diagnostic pipeline.

## Caching and failure handling

Repeated diagnostic calls can reuse cached results. `runHeteroTests()` builds
an analysis key from the model, data, selected tests, and relevant execution
parameters when caching is enabled and the optional hashing dependency is
available. Cached objects are returned using the same suite classes as newly
computed results.

Errors are analysed at the orchestration layer rather than being silently
discarded. The package records diagnostic context, can attempt a supported
fallback, and produces structured failure results for grouped analyses where
continuing with the remaining groups is preferable to aborting the entire
operation. For ordinary ungrouped calls, unrecoverable errors are still raised
to the caller.

## Memory-bounded auxiliary regressions

White, Breusch--Pagan, and Koenker diagnostics also have streaming
implementations. Their common computational problem is an auxiliary regression
whose statistic depends on its coefficient of determination. Rather than
constructing the complete auxiliary design matrix, the streaming code processes
observations in chunks and accumulates

\[
  X^\top X, \qquad X^\top y, \qquad
  \sum_i y_i, \qquad \text{and} \qquad \sum_i y_i^2.
\]

These quantities are sufficient to solve the auxiliary least-squares problem
and recover the residual and total sums of squares. If
\(\widehat{\beta}=(X^\top X)^{-1}X^\top y\), the residual sum of squares can
be written as

\[
  \mathrm{SSE}=y^\top y-\widehat{\beta}^\top X^\top y,
\]

while the total sum of squares is obtained from the accumulated first and
second moments of the response. The resulting \(R^2\) is therefore computed
without retaining all rows of the auxiliary design simultaneously.

This matters most for White's test, where squares and cross-products can expand
the auxiliary design substantially. The implementation also checks for
rank-deficient accumulated systems and can omit interaction expansion when the
number of regressors exceeds a configurable limit. If \\CRANpkg{Matrix} is
available, sufficiently large chunk cross-products may be evaluated using sparse
matrices.

`runHeteroTests()` estimates the size of the supplied data and uses these
streaming implementations automatically for supported diagnostics when the
configured threshold is exceeded. The threshold and chunk size remain user
controls, so the same API can be used for both ordinary in-memory and
memory-constrained analyses.

# Validation strategy

The validation strategy is organised around the claim being checked rather than
around the package function that happens to implement it. Three forms of
evidence are used: agreement with an established implementation, independent
reconstruction of a published statistic, and behavioural tests of statistical
or computational properties.

## Agreement with established implementations

Where another R implementation exposes the same statistic under the same
assumptions, the package test suite compares both the statistic and the
associated p-value. This is the strongest validation available for wrappers of
well-established methods because it checks not only formula transcription but
also practical details such as degrees of freedom, alternative hypotheses, and
centering conventions.

The current suite includes direct comparisons with \\CRANpkg{lmtest} for the
classical Breusch--Pagan, studentized Breusch--Pagan/Koenker, and
Goldfeld--Quandt tests [@zeileis2002lmtest]. Mean-centred Levene,
median-centred Brown--Forsythe, and Cook--Weisberg/non-constant-variance
diagnostics are checked against \\CRANpkg{car} [@fox2019car]. Group-wise
variance procedures are also compared with base R implementations for Bartlett
and Fligner--Killeen tests and, where available, with \\CRANpkg{vartest} for
Hartley and O'Brien procedures.

These comparisons use explicit numerical tolerances in the automated test
suite. The tolerance depends on the implementation pair: some functions match
to floating-point equality, while others are checked within a small numerical
tolerance because the two implementations may differ in intermediate linear
algebra without changing the intended statistic.

## Independent reconstruction

An external package is not always available, and using another implementation
is not sufficient when both could share the same mistake. For this reason, a
second group of tests reconstructs the claimed statistic directly from its
published definition or from lower-level base-R operations.

Examples include the Cook--Weisberg score statistic [@cook1983detection],
Harvey's multiplicative-variance LM statistic [@harvey1976estimating],
Szroeter's ordered-variance statistic, Engle's ARCH LM statistic
[@engle1982autoregressive], and the McLeod--Li procedure, which is checked
against a Ljung--Box calculation on squared residuals. The same approach is
used for O'Brien's transformed observations and for degrees-of-freedom
behaviour under rank-deficient auxiliary models.

This layer is particularly useful for implementation details that are easy to
get subtly wrong. Examples caught during package development include scaling of
ordered-variance statistics, auxiliary-model degrees of freedom, and the
construction of resampling null distributions.

## Behavioural and invariant checks

Some properties are better validated as invariants than by comparing a single
number. Property-based tests verify that supported diagnostics return valid
`htest` objects with finite statistics and p-values in ([0,1]) over randomly
generated inputs. Simulation-based regression tests check that selected
classical procedures have empirical Type-I error close to the nominal level,
using a tolerance based on the Monte Carlo standard error rather than a fixed
arbitrary margin.

Resampling procedures receive separate checks because an apparently plausible
implementation can still produce invalid p-values. Bootstrap and permutation
p-values use the finite-simulation convention

\[
  p = \frac{1 + \#\{T_b \ge T_{\mathrm{obs}}\}}
           {B_{\mathrm{eff}} + 1},
\]

which prevents an estimated p-value of exactly zero and accounts for failed
replicates through the effective number of simulations. Null-imposed residual
resampling is also checked for leverage correction, while unsupported model
classes or transformed responses are rejected rather than silently refitted
under a different statistical model.

Finally, the streaming implementations are compared directly with their
ordinary in-memory counterparts. For White, Breusch--Pagan, and Koenker tests,
the test suite checks equality of the statistic and p-value within numerical
tolerance for identical data. This verifies the principal claim of the
streaming layer: reduced peak memory use should not define a different
diagnostic.

## Reproducibility of the evidence

The article does not hard-code validation numbers from earlier package
versions. Reference comparisons and behavioural checks live in the package test
suite, while empirical size, power, and performance results reported later in
the paper are regenerated from the version analysed by the manuscript. This
separation is intentional. Unit and reference tests answer whether an
implementation is computing the intended object; simulation and benchmark
sections answer how those procedures behave under selected data-generating
processes and computational loads.

The distinction also limits the claims made here. Agreement with another
implementation is evidence of computational equivalence for the cases tested,
not proof that the underlying statistical procedure is appropriate for every
analysis. Likewise, satisfactory empirical size under a simulation design does
not establish finite-sample validity outside that design. The purpose of this
validation framework is therefore narrower: to make the software behaviour
auditable and to reduce the risk that integration work changes the statistic a
diagnostic is supposed to compute.

# Simulation study

The simulation study separates calibration under the null from power under
specific alternatives. This distinction is important because a procedure that
rejects frequently under a heteroscedastic alternative is not useful if it also
substantially exceeds its nominal rejection probability under homoscedasticity.

All experiments use the linear mean model

\[
  y_i = 1 + 2x_i + \sigma(x_i)\varepsilon_i,
  \qquad x_i \sim U(0,1),
\]

with independent errors. The significance level is \(\alpha=0.05\). Within
each Monte Carlo replication the same simulated data set is supplied to all
diagnostics, so differences in rejection rates are not inflated by using
different random samples for different tests.

The study includes White, classical Breusch--Pagan, Koenker, and a two-sided
Goldfeld--Quandt test ordered by \(x\). These procedures represent different
ways of detecting variance structure: general auxiliary-regression expansion,
linear score-type alternatives, a studentized variant less dependent on
normality, and an ordered two-sample variance comparison.

## Empirical size

Empirical size is estimated under \(\sigma(x)=1\) for sample sizes
\(n\in\{50,100,250\}\). Two error distributions are considered: Gaussian
errors and Student \(t_5\) errors rescaled to unit variance. The heavier-tailed
case is included because agreement with a reference implementation does not
imply that an asymptotic approximation is well calibrated under non-normal
finite samples.

For each reported rejection rate \(\widehat p\), the simulation records the
Monte Carlo standard error

\[
  \operatorname{MCSE}(\widehat p)
  = \sqrt{\frac{\widehat p(1-\widehat p)}{B_{\mathrm{eff}}}},
\]

where \(B_{\mathrm{eff}}\) is the number of valid replications. Approximate
95% Monte Carlo intervals are reported as
\(\widehat p \pm 1.96\operatorname{MCSE}(\widehat p)\), truncated to
\([0,1]\). These intervals quantify simulation uncertainty; they are not
confidence intervals for a population parameter estimated from observed data.

## Power alternatives

Power is studied under three variance patterns with Gaussian errors:

- **linear:** variance scale increases monotonically with \(x\);
- **step:** the variance changes abruptly at \(x=0.5\);
- **U-shaped:** variance is smallest near the centre of the covariate range and
  increases toward both extremes.

For each pattern, an effect-strength parameter
\(\gamma\in\{0.25,0.50,0.75,1.00\}\) interpolates between constant
variance and the target profile. The resulting standard-deviation vector is
renormalised to unit root-mean-square scale. Consequently, increasing
\(\gamma\) changes the *shape* of heteroscedasticity without simply changing
the overall noise magnitude. This avoids the common error of calling a global
multiplicative scale change an increase in heteroscedastic effect size.

The non-monotone U-shaped alternative is included deliberately. A test ordered
by a single covariate direction can be powerful against monotone changes while
having much less sensitivity to variance patterns that increase at both ends of
the covariate range. The purpose of the study is therefore comparative rather
than to identify a universally dominant diagnostic.

## Reproducible experiment

The complete experiment is implemented in
`paper/simulation_study.R`. The manuscript run uses an explicit seed and
writes both results and simulation metadata to `paper/generated/`:

```{r simulation-command, eval=FALSE}
system2(
  "Rscript",
  c("paper/simulation_study.R", "2000")
)
```

The default manuscript configuration uses 2,000 replications per design cell.
With a rejection probability near 0.05 this corresponds to a Monte Carlo
standard error of approximately 0.005, which is small enough to distinguish
material size distortions from ordinary simulation noise without pretending
that the estimated rejection rates are exact.

Numerical results and figures are intentionally read from the generated output
rather than copied from earlier drafts. A render therefore cannot silently
reuse the retired `power_summary.csv` values as evidence for the redesigned
experiment.

# Computational performance

Benchmarks will measure runtime and memory behaviour as sample size increases,
with particular attention to the streaming implementations. Results will be
regenerated for the release analysed in this article.

# A reproducible diagnostic workflow

This section will present one end-to-end analysis: fit a model, diagnose
non-constant variance, interpret conflicting diagnostics, apply an appropriate
remediation, and compare the resulting model diagnostics.

# Comparison with existing R software

The comparison will focus on software scope and workflow differences rather
than treating existing packages as competitors. In particular, the article
will distinguish specialised reference implementations from the unified
interface and orchestration provided by heteroTests.

# Discussion

No single heteroscedasticity diagnostic is uniformly preferable. Test choice
depends on the suspected variance structure, sample size, distributional
assumptions, dependence structure, and the inferential purpose of the model.
The package therefore exposes multiple procedures while retaining the
information needed to understand how each result was obtained.

# Summary

heteroTests provides a common software layer for heteroscedasticity diagnostics
in R. Its contribution is not a replacement for the underlying statistical
literature, but a reproducible and extensible framework that makes heterogeneous
diagnostics easier to validate, combine, scale, and integrate into modelling
workflows.
